---
title: "2nd assignment"
author: "Joan Prat Sicart"
date: "4/24/2019"
output: html_document
---

#PREPARE THE DATA

Import all the dataset get rid of the uninteresting columns and finally bind between them with their respective years 

```{r}
#import the datasets
marathon_2015 <- read.csv("marathon_results_2015.csv", header = TRUE, na.strings="", stringsAsFactors=FALSE)

marathon_2016 <- read.csv("marathon_results_2016.csv", header = TRUE, na.strings="", stringsAsFactors=FALSE)

marathon_2017 <- read.csv("marathon_results_2017.csv", header = TRUE, na.strings="", stringsAsFactors=FALSE)
#remember that = can be replaced by <-
# ()for function/ [object to found , colum] for a dataset (marathon$X5K is the same marathon[,"X5K"])

#exclude X1 from marathon marathon_2015 and marathon_2017 
marathon_2015 <-subset(marathon_2015, select = -c( X.1) )
marathon_2017 <-subset(marathon_2017, select = -c( X.1) )

```

Merge the three datasets into one called marathon, but before introduce the year as a colum to then be able to diferentiate among the diffrents marathons

```{r}
#add year colimn 
marathon_2015$Year = 2015
marathon_2016$Year = 2016
marathon_2017$Year = 2017

#merge the datasets
marathon = rbind(marathon_2015, marathon_2016, marathon_2017)

#delete the unnecessary tables
rm(marathon_2015, marathon_2016, marathon_2017)
```

Make sure that there aren't runners in the datasets without the time of one partial, becasuse it's considered that has cheated.

```{r}
#get the number of rows with all the partials completed 

#marathon =  subset( marathon,!marathon$X5K == "-" && !marathon$X10K== "-" && !marathon$X15K == "-" && !marathon$X20K== "-" && !marathon$X25K== "-" && !marathon$X30K== "-" && !marathon$X35K== "-" && !marathon$X40K== "-" || !marathon$Official.Time== "-")


```

```{r}
marathon = marathon[!marathon$X5K == "-",]
marathon = marathon[!marathon$X10K == "-",]
marathon = marathon[!marathon$X15K == "-",]
marathon = marathon[!marathon$X20K == "-",]
marathon = marathon[!marathon$Half == "-",]
marathon = marathon[!marathon$X25K == "-",]
marathon = marathon[!marathon$X30K == "-",]
marathon = marathon[!marathon$X35K == "-",]
marathon = marathon[!marathon$X40K == "-",]
marathon = marathon[!marathon$Official.Time == "-",]
```

Replace tha M value for 0 and F for 1, tranform the times to seconds, and countries to codes.

```{r}
#tranform M/F into numbers 
marathon$M.F[marathon$M.F == "M"]=0 
marathon$M.F[marathon$M.F == "F"]=1

#transform into seconds the times
library(lubridate)
marathon$X5K <- period_to_seconds(hms(marathon$X5K))
marathon$X10K <- period_to_seconds(hms(marathon$X10K))
marathon$X15K <- period_to_seconds(hms(marathon$X15K))
marathon$X20K <- period_to_seconds(hms(marathon$X20K))
marathon$X25K <- period_to_seconds(hms(marathon$X25K))
marathon$X30K <- period_to_seconds(hms(marathon$X30K))
marathon$X35K <- period_to_seconds(hms(marathon$X35K))
marathon$X40K <- period_to_seconds(hms(marathon$X40K))
marathon$Official.Time <- period_to_seconds(hms(marathon$Official.Time))

#Convert the contries in numbers
library(countrycode)
library(plyr)
marathon$Country <- countrycode(marathon[, "Country"], "ioc", "genc3n", nomatch = NULL)
marathon$Country <- mapvalues(marathon[, "Country"], from=c("FLK", "SIN", "TCA", "TRI", "TWN"), to=c("246", "702", "796", "780", "158")) #because the function doesn't convert those types

```

We will introduce the temperature variable in function with the day that had been runned the marathon, to do so, as we don't have specifics range of hours we will introduce the average temperature in Fº for each day, Therefore, for the 2015 year the average temperature was 47, for the 2016 year was 55, and finally, for the 2017 year was 61

```{r}

# Add temperature as a variable
marathon [marathon$Year == 2015, "Temperature"]=46 
marathon [marathon$Year == 2016, "Temperature"]=55 
marathon [marathon$Year == 2017,"Temperature"]=62 

```

To simplify also the things we will exclude also the variables without interest

```{r}
#exclude the variables without interest
marathon <-subset(marathon, select = -c(X, Half, Overall, Gender, Pace, City, State, Name, Proj.Time, Citizen) )
```

As we've considered that don't have the same behaviour the all the runners, for instance the elite group follow a continuos pace during the race while the amateurs their pace change during the race, we are going to select the elite group and work with it.

In comparison with the female gender, for the male gender there isn't any indicator to distinguish among elite and amateur runners, so we'll get the percentatge of female runners and assume that there is the same percentage for the male.

```{r}
#Make a table with the elite female

marathon_elite_female = marathon[marathon$M.F == 1 & substr(marathon$Bib, start = 1, stop = 1) == "F" & (marathon$Age >= 20 & marathon$Age <= 40),]

# Get the percentaje of elite female runners

female_elite = sum(marathon$M.F == 1 & substr(marathon$Bib, start = 1, stop = 1) == "F")
female_amateur = sum(marathon$M.F == 1 & substr(marathon$Bib, start = 1, stop = 1) != "F")
percentage_elite = female_elite / (female_elite+female_amateur)

#Make a table with the elite male considering thta the percentatge among elite and amateur is the same as the female runners

marathon_elite_male = marathon[marathon$M.F == 0 & (marathon$Age >= 20 & marathon$Age <= 40),]
# sort them in function with the final time
marathon_elite_male = marathon_elite_male[order(marathon_elite_male$Official.Time),] 
# get the first X% of male runners
marathon_elite_male = head(marathon_elite_male, floor(nrow(marathon_elite_male)*percentage_elite)) 
#Join the two tales (elite_male, elite_female)
marathon_elite = rbind(marathon_elite_female, marathon_elite_male) 
#remove unncessary tables
rm(marathon_elite_female, marathon_elite_male)
```

make sure that all the values are integers and there isn't rows with one of the value null

```{r}

marathon_elite <-subset(marathon_elite, select = -c(Bib) )

#Make sure that all the values are numeric
character_vars <- lapply(marathon_elite, class) == "character"
marathon_elite[, character_vars] <- lapply(marathon_elite[, character_vars], as.numeric)

#Make sure that there aren't rows with a null value
#marathon_elite <- data.frame(t(na.omit(t(marathon_elite)))) 
#// to omit those columns that have null values to avoid problems with the null (in partiuclar are Country  and Bib) but is better to delete those rows that have that fields null.
marathon_elite <- na.omit(marathon_elite)
```

#PCA

execute the Principal Component Analisis in order to visualize the different relations in the dataset with the variables already choosen.

```{r}
# PCA

marathon_elite.pca <- prcomp(marathon_elite, center = TRUE,scale. = TRUE) # center and scale the data
summary(marathon_elite.pca)

# Plot the PCA
library(ggfortify)
autoplot(marathon_elite.pca, data = marathon_elite, loadings = TRUE, loadings.label = TRUE)
```

Observing the PCA, we can conclude that, as it was normal, the partial times are strongly related with the other partial times, the official time, the bib and the ranking of the runners, however, that kind of information won't be given to us when modelling the behaviour of a runner, since that are dates that aren't known previously of the simulation. Looking to the ohter variables we can see that the factors that have more impact on the partial results are the gender and the age, but temperature don't have a strong relation with the first partial for instance, however we are going to check how is better the model, with the temperature or without it.


#LINEAR MODEL

Linear model for the first partial X5K

```{r}
l_model = lm(X5K ~ M.F + Age + Temperature + Country, data = marathon_elite)
summary(l_model)


#As we have already annunced previously looking in the PCA Temperature don't have  a strong relation with the first partial times since the p-value is greater than 0.05, which is logical, since that in the first km don't influence too much the weather, therefore, it's necessary make another model without the Temperature


#0 - 5Km linear model for all the marathon 
l_model = lm(X5K ~ M.F + Country , data = marathon_elite)
summary(l_model)
```

Linear model for the second partial 

```{r}
marathon_elite$X5K_X10K = marathon_elite$X10K - marathon_elite$X5K
l_model = lm(X5K_X10K ~ M.F + Age + Temperature + Country, data = marathon_elite)
summary(l_model)

# 5-10 km final linear model
l_model = lm(X5K_X10K ~ M.F + Country , data = marathon_elite)
summary(l_model)

```

Linear model for the third partial 

```{r}
marathon_elite$X10K_X15K = marathon_elite$X15K - marathon_elite$X10K 
l_model = lm(X10K_X15K ~ M.F + Age + Temperature + Country, data = marathon_elite)
summary(l_model)

# 10-15 km final linear model
l_model = lm(X10K_X15K ~ M.F + Temperature + Country, data = marathon_elite)
summary(l_model)

```

Linear model for the fourth partial

```{r}
marathon_elite$X15K_X20K = marathon_elite$X20K - marathon_elite$X15K 
l_model = lm(X15K_X20K ~ M.F + Age + Temperature + Country, data = marathon_elite)
summary(l_model)

# 15 - 20 km linear model 
l_model = lm(X15K_X20K ~ M.F + Country, data = marathon_elite)
summary(l_model)

```

Linear model for the fifth partial

```{r}
marathon_elite$X20K_X25K = marathon_elite$X25K - marathon_elite$X20K 
l_model = lm(X20K_X25K ~ M.F + Age + Temperature + Country, data = marathon_elite)
summary(l_model)

# 20 - 25 km linear model 
l_model = lm(X20K_X25K ~ M.F + Country, data = marathon_elite)
summary(l_model)
```

Linear model for the sixth partial

```{r}
marathon_elite$X25K_X30K = marathon_elite$X30K - marathon_elite$X25K 
l_model = lm(X25K_X30K ~ M.F + Age + Temperature + Country, data = marathon_elite)
summary(l_model)

# 25 - 30 km linear model 
l_model = lm(X25K_X30K ~ M.F + Country, data = marathon_elite)
summary(l_model)

```

Linear model for the seventh partial

```{r}
marathon_elite$X30K_X35K = marathon_elite$X35K - marathon_elite$X30K 
l_model = lm(X30K_X35K ~ M.F + Age + Temperature + Country, data = marathon_elite)
summary(l_model)

# 30 - 35 km linear model
l_model = lm(X30K_X35K ~ M.F + Country, data = marathon_elite)
summary(l_model)

```


Linear model for the eighth partial

```{r}
marathon_elite$X35K_X40K = marathon_elite$X40K - marathon_elite$X35K 
l_model = lm(X35K_X40K ~ M.F + Age + Temperature + Country, data = marathon_elite)
summary(l_model)

# 35 - 40 km linear model 
l_model = lm(X35K_X40K ~ M.F  + Country, data = marathon_elite)
summary(l_model)

```

Linear model for the ninenth partial

```{r}
marathon_elite$X40K_Final = marathon_elite$Official.Time - marathon_elite$X40K 
l_model = lm(X40K_Final ~ M.F + Age + Temperature + Country, data = marathon_elite)
summary(l_model)

# 40 - Final km linear model
l_model = lm(X40K_Final ~ M.F  + Country , data = marathon_elite)
summary(l_model)
```

As we can see in all the final models all the factors have a low p-value, even the intercept, which means that all the values in the linear model are important when modelling the behavior of the partial times, moreover, the adusjtest r square has also a good value and therefore the model is quite accurate.

#VALIDE THE LINEAR MODELS OBTAINED

Finally we'll be necessary validate the linear models by checking the homoscedasticity, the independence, and the normal distribution of the data 

```{r}
# Test Linear Model hypothesis
#Durbin Watson (independent)
library("lmtest")
dwtest(l_model, alternative ="two.sided")
#Shapiro test (normal)
shapiro.test(residuals(l_model))
#Breusch Pagan test (homoscedasticity)
lmtest::bptest(l_model)

```

!!!!!!!!!!!!!!!!!!Tant al durbin com el shapiro test estan a la l'hipotesis alternativa, això significa que ni les dades son independents ni segueixen una distribució normal
(Since the p-value for the normality assumption (Shapiro-Wilk) is lower than 0,05, we must reject the assumption that the populations from which samples are selected are normal. Therefore, this linear model should not be used to predict the times at 10km.)

#COMPROVE EH REGARDING THE PACE OF AN AMATEUR RUNNER AND AN ELITE RUNNER

In order to check how the level of professionality of each runner affect in the pace, first of all, we are going to divide the dataset into two datasets whether they are elite or amteur and  the with each dataset construct a table where the partial times fo each runners are associated with the distance runned, thus, then, making a linear regression we'll be able to see if as we think the elite runners keep a similar pace and the amateur slowly decrease the pace or not


```{r}

#female elite runners
marathon_female_1=marathon[marathon$M.F == "1" & substr(marathon$Bib, start = 1, stop = 1) == "F",]
#female amateur runners
marathon_female_2=marathon[marathon$M.F == "1" & substr(marathon$Bib, start = 1, stop = 1) != "F",]

#Keep only the partial times for each Group
marathon_female_1 = subset(marathon_female_1, select=c(X5K, X10K, X15K, X20K, X25K, X30K, X35K, X40K))
marathon_female_2 = subset(marathon_female_2, select=c(X5K, X10K, X15K, X20K, X25K, X30K, X35K, X40K))

#Table for the marathon elite runners with the distance and the times
len = nrow(marathon_female_1) 
final_table_female_1 = data.frame(Distance_km = rep(5, len), Time_seconds = marathon_female_1$X5K)
j=2
#if we don't initialitze and then bind it foreach distance the we are going to retrieve just the results for the las 40km 
for (i in seq(10, 40, by=5)){
  table_female_1= data.frame(Distance_km = rep(i, len), Time_seconds = marathon_female_1[j])
  names(table_female_1)[1] = "Distance_km"
  names(table_female_1)[2] = "Time_seconds"
  final_table_female_1 = rbind(final_table_female_1, table_female_1)
  j=j+1
}

#Table for the marathon amateur runners with the distance and the times

len = nrow(marathon_female_2) 
final_table_female_2 = data.frame(Distance_km = rep(5, len), Time_seconds = marathon_female_2$X5K)
j=2
for (i in seq(10, 40, by=5)){ 
  len = nrow(marathon_female_2)
  table_female_2 = data.frame(Distance_km = rep(i, len), Time_seconds = marathon_female_2[j])
  names(table_female_2)[1] = "Distance_km"
  names(table_female_2)[2] = "Time_seconds"
  final_table_female_2 = rbind(final_table_female_2,table_female_2)
  j=j+1
}

```


Check with Kolmogorov-Smirnov if follows a normal distribution or not

```{r}
#ks.test(x, y, ..., alternative = c("two.sided", "less", "greater"), exact = NULL, tol=1e-8, simulate.p.value=FALSE, B=2000)

#does x and y come from the same distribution?
#ks.test(x, y)

#spark.kstest(marathon, testCol = "test",
 # nullHypothesis = c("norm"), distParams = c(0, 1))

hist(marathon_elite$Official.Time)
```


Linear models for each female table

```{r}
#linear regresions
summary(lm(Time_seconds ~ Distance_km, data=final_table_female_1))
summary(lm(Time_seconds ~ Distance_km, data=final_table_female_2))

#plots of the data
boxplot(Time_seconds ~ Distance_km, data=final_table_female_1, col=rainbow(7), xlab='Distance [km]', ylab='Time [s]')
boxplot(Time_seconds ~ Distance_km, data=final_table_female_2, col=rainbow(7), xlab='Distance [km]', ylab='Time [s]')
```

#CHECK IF FOR EACH INDEPENDENT VARIABLE THERE ARE DIFFERENT NORMAL DISTRIBUTION

```{r}
# ANOVA for the Temperature
anova_model = aov(Official.Time ~ Temperature, data=marathon)
summary(anova_model) 
boxplot(Official.Time ~ Temperature, data=marathon, col=rainbow(7), ylab='Time ', xlab='Temperature ')
```

```{r}
# ANOVA for the Age
marathon$Old[marathon$Age <= 40] = 0
marathon$Old[marathon$Age > 40] = 1
anova_model = aov(Official.Time ~ Old, data=marathon)
summary(anova_model) 
boxplot(Official.Time ~ Old, data=marathon, col=rainbow(7), xlab='Age', ylab='Time ', names = c("<40", ">40"))
```

```{r}
# ANOVA for the Gender
anova_model = aov(Official.Time ~ M.F, data=marathon)
summary(anova_model) 
boxplot(Official.Time ~ M.F, data=marathon, col=rainbow(7), xlab='Gender', ylab='Time', names = c("Male", "Female"))
```

```{r}
# ANOVA for the Country
marathon$African[marathon$Country >= 230 && marathon$Country <= 330] = 1
marathon$African[marathon$Country <= 230 && marathon$Country >= 330] = 0
anova_model = aov(Official.Time ~ African, data=marathon)
summary(anova_model) 
boxplot(Official.Time ~ M.F, data=marathon, col=rainbow(7), xlab='Nationality', ylab='Time', names = c("African", "Not African"))
```

#GET THE MEANS OF THE RESPONSES OF THE SYSTEMS FOR THE DIFFERENT FACTORS

First of all we are going to divide the responses of the system in function of the values of their variables, and then get the mean foreach scenario for the elite dataset, the one we've studied during the project

```{r}

#including the the country variable where - are from all the countries and + are from the african country for example

marathon_elite$Scenario[marathon_elite$Country <= 220 | marathon_elite$Country >= 380 & marathon_elite$M.F == 0 & marathon_elite$Age < 30 & marathon_elite$Temperature <54 ] = 1
marathon_elite$Scenario[marathon_elite$Country <= 220 | marathon_elite$Country >= 380 & marathon_elite$M.F == 0 & marathon_elite$Age < 30 & marathon_elite$Temperature >= 54] = 2
marathon_elite$Scenario[marathon_elite$Country <= 220 | marathon_elite$Country >= 380 & marathon_elite$M.F == 0 & marathon_elite$Age >= 30 & marathon_elite$Temperature <54] = 3
marathon_elite$Scenario[marathon_elite$Country <= 220 | marathon_elite$Country >= 380 & marathon_elite$M.F == 0 & marathon_elite$Age >= 30 & marathon_elite$Temperature >= 54] = 4
marathon_elite$Scenario[marathon_elite$Country <= 220 | marathon_elite$Country >= 380 & marathon_elite$M.F == 1 & marathon_elite$Age < 30 & marathon_elite$Temperature <54] = 5
marathon_elite$Scenario[marathon_elite$Country <= 220 | marathon_elite$Country >= 380 & marathon_elite$M.F == 1 & marathon_elite$Age < 30 & marathon_elite$Temperature >= 54] = 6
marathon_elite$Scenario[marathon_elite$Country <= 220 | marathon_elite$Country >= 380 & marathon_elite$M.F == 1 & marathon_elite$Age >= 30 & marathon_elite$Temperature <54] = 7
marathon_elite$Scenario[marathon_elite$Country <= 220 | marathon_elite$Country >= 380 & marathon_elite$M.F == 1 & marathon_elite$Age >= 30 & marathon_elite$Temperature >= 54] = 8
marathon_elite$Scenario[marathon_elite$Country > 220 & marathon_elite$Country <= 380 & marathon_elite$M.F == 0 & marathon_elite$Age < 30 & marathon_elite$Temperature <54 ] = 9
marathon_elite$Scenario[marathon_elite$Country > 220 & marathon_elite$Country <= 380 & marathon_elite$M.F == 0 & marathon_elite$Age < 30 & marathon_elite$Temperature >= 54] = 10
marathon_elite$Scenario[marathon_elite$Country > 220 & marathon_elite$Country <= 380 & marathon_elite$M.F == 0 & marathon_elite$Age >= 30 & marathon_elite$Temperature <54] = 11
marathon_elite$Scenario[marathon_elite$Country > 220 & marathon_elite$Country <= 380 & marathon_elite$M.F == 0 & marathon_elite$Age >= 30 & marathon_elite$Temperature >= 54] = 12
marathon_elite$Scenario[marathon_elite$Country > 220 & marathon_elite$Country <= 380 & marathon_elite$M.F == 1 & marathon_elite$Age < 30 & marathon_elite$Temperature <54] = 13
marathon_elite$Scenario[marathon_elite$Country > 220 & marathon_elite$Country <= 380 & marathon_elite$M.F == 1 & marathon_elite$Age < 30 & marathon_elite$Temperature >= 54] = 14
marathon_elite$Scenario[marathon_elite$Country > 220 & marathon_elite$Country <= 380 & marathon_elite$M.F == 1 & marathon_elite$Age >= 30 & marathon_elite$Temperature <54] = 15
#l'únic valor the country per aquest scenario es 404 i si amplio fins aquest valor el parametre country llavors es menja el scenario 7, per tant direm que tant l'escenari 7 com el 15 tenen el mateix output
marathon_elite$Scenario[marathon_elite$Country > 220 & marathon_elite$Country <= 405 & marathon_elite$M.F == 1 & marathon_elite$Age >= 30 & marathon_elite$Temperature >= 54] = 16

# Get the mean for each Scenario
boxplot(Official.Time ~ Scenario, data=marathon_elite, col=rainbow(24), xlab='Scenario', ylab='Time')
means_real_system = aggregate(Official.Time ~ Scenario, data=marathon_elite, mean)


```